---
title: Automatically assigning celltypes to scChIC-seq data from bulk ChIP-seq
author: Jake Yeung
path: "~/projects/scchic/scripts/scripts_analysis/revisions/probabilistic_celltype_assignment.R"
date: 2019-09-14
output:
   html_document:
     toc: true
     highlight: zenburn
---

## Introduction

I mentioned before that I think the scChIC-seq data (and probably scRNA-seq) of a single cell contains a lot more information than people think, 
because we’re able to do things like unmix doubly stained cells on a cell-by-cell basis, if we define the right likelihood model.

The nice thing about the bone marrow is that there is publicly available chip-seq data from which to compare (namingly, Lara-Astiaso, Science 2014 from Amit lab). 

A naive way of assigning celltypes is to make a 2D summary of your data, create clusters, generate pseudobulk from clusters, and find which pseudobulk corresponds to which celltype, 
given a publicly available dataset (we did this by just calculating correlations, between pseudobulk and public data).

The output of this was less than satisfactory, mostly because it's difficult to compare across many correlations 
(interpretation of correlation nonlinear and complicated: correlation of 0.7 versus 0.6 is not the same as correlation of 0.9 versus 0.8. Difference between 0.9 and 0.8 is much bigger than 0.7 versus 0.6.).

It would be nice to systematically calculate the likelihood that the single-cell data was generated from a particular ChIP-seq bulk sample, then compare across many bulk samples to 
calculate the probability that a cell came from a particular sample. 

The main take home message is that there is a surprisingly large amount of information coming from scChIC-seq data of an individual cell such that you can do celltype calling without creating
pseudobulk (as long as you have good collection of annotated bulk data). I am pretty sure this also applies to scRNA-seq. 



## The generative model of scChIC-seq data

We assume the single scChIC-seq data is being generated from a multinomial process (e.g., drawing balls in an urn). 
In the balls in and urn analogy, the multinomial process is parametrized by a vector $\vec{p}$ of length $|p| = K$ representing the probability of drawing a ball from each of the K colors.
In the scChIC-seq data, drawing a ball is sequencing a UMI, and the K colors represent the K different genomic regions from which to draw a UMI. 

The multinomial likelihood $L$ of seeing a vector $\vec{y}$ of UMI counts across the K genomic bins in a cell (total UMI counts in the cell, $N = \sum_i{y}$), is:
$$L = P(\vec{y} | \vec{p}, N) \propto \prod_{k=1}^K{\left( p_k \right) ^ {y_k}}$$

If you have M annotated celltypes from which to select, then the game is to parametrize M probability vectors $\vec{p_1}, \vec{p_2}, ..., \vec{p_M}$ such that you can take your single-cell observation $\vec{y}$ and
calculate the likelihood for each of the probability vectors, i.e., calculate $L_1, L_2, ..., L_M$ and then select the model with the highest likelihood. We can therefore use $\vec{L}$ to calculate the probability
that a cell came a celltype.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

library(JFuncs)
library(dplyr)
library(ggplot2)
library(data.table)
library(topicmodels)
library(tidytext)
library(umap)
library(ggrepel)

library(tidyr)

library(hash)
library(igraph)

library(ChIPseeker)
library(GenomicRanges)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

library(scchicFuncs)
library(ggrepel)

library(preprocessCore)

library(here)

setwd(here())

source("scripts/Rfunctions/VariabilityFunctions.R")
source("scripts/Rfunctions/Aux.R")
source("scripts/Rfunctions/AuxB6.R")
source("scripts/Rfunctions/PlotFunctions.R")

# load publicly available data
indir <- "/Users/yeung/data/scchic/from_cluster/revision_objs/public_data_objs"

infs <- list.files(indir, pattern = "*.tab", full.names = TRUE)
ctypes <- sapply(infs, function(inf) strsplit(basename(inf), split = "_")[[1]][[2]], USE.NAMES = FALSE)
names(infs) <- ctypes

dat.chipseq <- lapply(ctypes, function(ctype){
  dat.tmp <- fread(infs[[ctype]])
  dat.tmp$ctype <- ctype
  # take first 4 columns for now, rest are chic
  dat.tmp <- dat.tmp[, c(1:4)]
  colnames(dat.tmp) <- c("chr", "start", "end", ctype)
  return(dat.tmp)
}) %>%
  purrr::reduce(left_join)
# https://stackoverflow.com/questions/8091303/simultaneously-merge-multiple-data-frames-in-a-list/34393416#34393416

# remove lowly expressed bins and do PCA
jcutoff <- 5
cols.keep <- !colnames(dat.chipseq) %in% c("chr", "start", "end")
dat.filt <- dat.chipseq[which(rowMeans(dat.chipseq[, cols.keep]) > jcutoff), ]


# load all chicseq data
jmark <- "H3K4me1"

# load trajs, load H3K4me3 separately because we analyzed it separately
inf.traj.stringent <- "/Users/yeung/data/scchic/robjs/B6_objs_stringent/traj_objs_H3K4me3.stringent.Rdata"
load(inf.traj.stringent, v=T)
dat.umap.long.trajs.stringent <- dat.umap.long.trajs$H3K4me3
trajs.stringent <- trajs$H3K4me3
trajs.objs.stringent <- trajs.objs$H3K4me3
inf.traj <- "/Users/yeung/data/scchic/robjs/B6_objs/traj_objs_all_marks.Rdata"
load(inf.traj, v=T)
dat.umap.long.trajs$H3K4me3 <- dat.umap.long.trajs.stringent
trajs$H3K4me3 <- trajs.stringent
trajs.objs$H3K4me3 <- trajs.objs.stringent


# get probability
# # load data from LDA
indir <- "/Users/yeung/data/scchic/tables/bamlist_for_merging_build95_B6"
jmarks <- c("H3K4me1", "H3K4me3", "H3K27me3", "H3K9me3"); names(jmarks) <- jmarks
infs <- lapply(jmarks, function(jmark) list.files(indir, pattern = paste0(jmark, ".RData"), full.names = TRUE))
infs.stringent <- "/Users/yeung/data/scchic/tables/bamlist_for_merging_build95_B6_stringent/dat_umap_long_with_louvain.H3K4me3.stringent_filter.RData"
infs$H3K4me3 <- infs.stringent
tm.result.lst <- lapply(infs, LoadGetTmResult)


inf.raw <- paste0("/Users/yeung/data/scchic/count_mat_binfilt_cellfilt_for_LDA/B6_", jmark, "_pcutoff_0.95_binfilt_cellfilt.2019-05-11.RData")
load(inf.raw, v=T)



# keep only bins that are interesting across topics

term.mat <- tm.result.lst[[jmark]]$terms

eryth.topic <- 7
lymph.topic <- 14
lymph.topic2 <- 28
granu.topic <- 47
granu.topic2 <- 3
nk.topic <- 40
mf.topic <- 13

topics.keep <- c(7,14,28,47,3,40,13)
names(topics.keep) <- topics.keep

nterms.keep <- 1000
terms.keep <- lapply(topics.keep, FUN = function(jtop){
  eryth.terms <- term.mat[jtop, ]
  jterms <- names(head(sort(eryth.terms, decreasing=TRUE), n = nterms.keep))
  return(jterms)
}) %>%
  unlist()

# hande duplicates
terms.keep <- unique(terms.keep)



```


## Analysis and results

### Quick PCA of the Lara-Astiaso ChIP-seq bone marrow data

Check that the ChIPseq data looks reasonable.

```{r pcacheck}

dat.input <- as.matrix(t(subset(dat.filt, select = c(-chr, -start, -end))))
dat.input.norm <- t(scale(t(dat.input), center = TRUE, scale = TRUE))
dat.input.norm <- scale(dat.input.norm, center = TRUE, scale = TRUE)

dat.pca <- prcomp(dat.input.norm, center = FALSE, scale. = FALSE)

dat.proj <- dat.input %*% dat.pca$rotation %*% diag(dat.pca$sdev)

dat.proj.long <- data.frame(celltype = rownames(dat.proj), pc1 = dat.proj[, 1], pc2 = dat.proj[, 2], stringsAsFactors = FALSE)

ggplot(dat.proj.long, aes(x = pc1, y = pc2, label = celltype)) + geom_point() + geom_text_repel() + 
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  geom_vline(xintercept = 0) + geom_hline(yintercept = 0)

```

### Normalize across samples to ensure ChIP-seq signal across bulk samples is comparable

We will use quantlile normalization to ensure comoparability across samples. 
Since we don't care about the absolute value of the underlying signal, we normalize the quantiles
so samples are comparable. 

The key is you want relative signal across bins to be comparable across samples. I think quantile
normalization is an easy way to achieve this (I tried without, still works for this dataset, but doing the
quantile normalization is likely to be more robust).


```{r normalizemagic}

# handle the NaNs
dat.mat <- as.matrix(dat.chipseq[, -c(1,2,3)])
rownames(dat.mat) <- paste(dat.chipseq$chr, paste(dat.chipseq$start, dat.chipseq$end, sep = "-"), sep = ":")

# set NaNs to zeros (rowwise??)
dat.mat[which(is.nan(dat.mat))] <- 0  # we can't have zeros in probability vector, fill these zeros with something else later

# remove the top outlier in each
for (col.i in seq(ncol(dat.mat))){
  print(col.i)
  vec.tmp <- dat.mat[, col.i]
  row.i <- which(vec.tmp == max(vec.tmp))
  dat.mat[row.i, col.i] <- 0
}

# boxplots can vary widely across samples
boxplot(dat.mat, main = "Before normalization")


```


## Set up probability vectors for multinomial likelihoods

Now that we have processed our reference dataset, let's construct our probability vectors and calculate likelihoods.

```{r probs}


dat.topics.filt <- dat.mat[unlist(terms.keep), ]

# handle zeros
zero.fill <- min(dat.topics.filt[which(dat.topics.filt > 0)])
dat.topics.filt[which(dat.topics.filt == 0)] <- zero.fill
dat.topics.filt <- sweep(dat.topics.filt, MARGIN = 2, STATS = colSums(dat.topics.filt), FUN = "/")

cnames.tmp <- colnames(dat.topics.filt)
rnames.tmp <- rownames(dat.topics.filt)
dat.topics.filt <- preprocessCore::normalize.quantiles(dat.topics.filt, copy = TRUE)  # strong normalization,
colnames(dat.topics.filt) <- cnames.tmp
rownames(dat.topics.filt) <- rnames.tmp

boxplot(dat.topics.filt, main = "Data used for constructing probabilities")
probs.lst.filt <- as.list(as.data.frame(dat.topics.filt))
# name the list just to be safe
probs.lst.filt <- lapply(probs.lst.filt, function(x){
  names(x) <- rownames(dat.topics.filt)
  return(x)
}) 


# use same bins between ChIC and ChIP

# dat.mat.filt <- dat.mat
bins.keep <- rownames(dat.mat)
# filter for bins 
# add chr to rownames
rownames(count.dat$counts) <- sapply(rownames(count.dat$counts), function(x) paste0("chr", x), USE.NAMES = FALSE)
count.filt <- count.dat$counts[bins.keep, ]




# check we didnt' filter out too many bins
par(mfrow=c(2,1), mar=c(5.1, 4.1, 4.1, 2.1), mgp=c(3, 1, 0), las=0)
plot(hist(Matrix::colSums(count.filt), breaks = 100))
plot(hist(Matrix::colSums(count.dat$counts), breaks = 100))
par(mfrow=c(1,1), mar=c(5.1, 4.1, 4.1, 2.1), mgp=c(3, 1, 0), las=0)

```

## Calculate likelihood cell by cell


```{r infercelltype}

all.cells <- subset(dat.umap.long.trajs[[jmark]])$cell
names(all.cells) <- all.cells
cell.names <- names(all.cells)

cell.counts <- Matrix::colSums(count.filt)  # note we keep sliding window
cell.counts.downsamp <- Matrix::colSums(count.filt)

# calculate data given model 
topics.keep <- c(7,14,28,47,3,40,13)
names(topics.keep) <- topics.keep

nterms.keep <- 1000
terms.keep <- lapply(topics.keep, FUN = function(jtop){
  eryth.terms <- term.mat[jtop, ]
  jterms <- names(head(sort(eryth.terms, decreasing=TRUE), n = nterms.keep))
  return(jterms)
}) %>%
  unlist()

# hande duplicates
terms.keep <- unique(terms.keep)

LL.ctype.lst <- lapply(all.cells, function(cell.name){
  cell.vec <- count.filt[terms.keep, cell.name]
  LL.vec <- sapply(probs.lst.filt, function(jprob){
    assertthat::assert_that(all(names(cell.vec) == names(jprob)))
    return(dmultinom(x = cell.vec, prob = jprob, log = TRUE))
  })
})
# calculate probability of model given data 
p.ctype.lst <- lapply(LL.ctype.lst, SoftMax)

# summaize
LL.dat <- lapply(cell.names, function(cname){
  LL.vec <- LL.ctype.lst[[cname]]
  p.vec <- p.ctype.lst[[cname]]
  cell.count = cell.counts[[cname]]
  cell.count.downsamp = cell.counts.downsamp[[cname]]
  if (all(is.infinite(LL.vec))){
    LL.max <- NA
    p.max <- NA
    best.ctype <- NA
  } else {
    LL.max <- max(LL.vec)
    p.max <- max(p.vec)
    best.ctype <- names(which.max(LL.vec))
  }
  dat.tmp <- data.frame(cell = cname, LL.max = LL.max, p.max = p.max, ctype.pred = best.ctype, cell.size = cell.count, cell.count.downsamp = cell.count.downsamp)
  return(dat.tmp) 
}) %>%
  bind_rows()

LL.sum <- LL.dat %>%
  group_by(ctype.pred) %>%
  summarise(ncell = length(ctype.pred))

print(LL.sum)

LL.dat.merge <- left_join(dat.umap.long.trajs[[jmark]], LL.dat)

```

### Show UMAP of celltype assignments


```{r output}

cbPalette <- c("#696969", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#006400", "#FFB6C1", "#32CD32", "#0b1b7f", "#ff9f7d", "#eb9d01", "#7fbedf")

# reduce number of colors by merging known groups together
# collapse EryA and EryB together
# collapse CD4, CD8 together
LL.dat.merge.pretty <- LL.dat.merge %>%
  rowwise() %>%
  mutate(ctype.pred = ifelse(ctype.pred == "EryA" | ctype.pred == "EryB", "EryAorB", ctype.pred),
         ctype.pred = ifelse(ctype.pred == "CD4" | ctype.pred == "CD8", "CD4or8", ctype.pred))

m.pretty <- ggplot(LL.dat.merge.pretty, 
       aes(x = umap1, y = umap2, color = ctype.pred)) + geom_point(alpha=0.75, size = 3) + 
  scale_color_manual(values = cbPalette) +
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  ggtitle(jmark, "Probabilistic celltype inference using multinomial likelihood, \n with Ido Amit ChIP-seq data as underlying genomic region proportions")
print(m.pretty)

# show cell by cell

# regress out cell.size relationship with LL

ggplot(LL.dat.merge, aes(x = log10(cell.size), y = LL.max)) + geom_point() + 
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())

ggplot(LL.dat.merge, aes(x = log10(cell.size), y = -LL.max)) + geom_point() + 
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())

ggplot(LL.dat.merge, aes(x = cell.size, y = -LL.max)) + geom_point() + 
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())

jfit <- lm(formula = LL.max ~ cell.size + I(sqrt(cell.size)), data = LL.dat.merge)
# jfit <- lm(formula = LL.max ~ cell.size + I(log(cell.size)), data = LL.dat.merge)
plot(LL.dat.merge$cell.size, LL.dat.merge$LL.max, pch = 20)
lines(sort(LL.dat.merge$cell.size), predict(jfit, newdata = data.frame(cell.size = sort(LL.dat.merge$cell.size))), type = "l", lwd = 5, col = "blue")

# get corectted LL.maxx
LL.dat.merge.cor <- LL.dat.merge %>%
  ungroup() %>%
  mutate(LL.pred = predict(jfit)) %>%
  rowwise() %>%
  mutate(LL.max.norm = LL.max - LL.pred)

# when cell size is small, you can't get large deltas across LL. 
# when cell size is big, then things start becoming very unlikely, therefore also less deltas? Strange
ggplot(LL.dat.merge.cor, aes(x = log10(cell.size), y = LL.max.norm)) + geom_point() +
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# comparing likelihoods across cells still not solved!
m.likelihoods <- ggplot(LL.dat.merge.cor %>% mutate(ctype.pred = ifelse(is.na(ctype.pred), "znone", ctype.pred)), 
       aes(x = umap1, y = umap2, color = LL.max)) + geom_point() + 
  scale_color_viridis_c() + theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  facet_wrap(~ctype.pred)
print(m.likelihoods)
```


## Discussion

Inferring cell type cell-by-cell is possible because there is a lot of information from individual cells of scChIC-seq data (no clustering needing). 

Of course, if there are novel celltypes that are not in the model, the method does not figure this out. We would need to interrogate the likelihoods of each cell and 
find outlier cells that seem to be very unlikely across all models. 



