---
title: 'Comparing '
author: "Jake Yeung"
date: "2019-01-02"
output: 
  html_document
---

# Introduction
```{r init, echo=TRUE, warning=FALSE, message=TRUE}
knitr::opts_chunk$set(fig.width=18, fig.height=10,
                      echo=TRUE, warning=FALSE, message=TRUE)
setwd("~/projects/scChiC")
source("scripts/Rfunctions/FitFunctions.R")
library(dplyr)
library(ggplot2)
library(hash)
library(umap)
# library(destiny)
library(princurve)
# load count matrices, LDA output, pseudotimes, fits

load("~/projects/scChiC/notebooks/notebook_objs/BM-H3K27me3.AvO_filt.Robj", v=T)
load("~/projects/scChiC/notebooks/notebook_objs/lda_out.meanfilt.K-12.Robj", v=T)
load("~/projects/scChiC/outputs_R/fit_output/glm_fits_100kb_windows_withOffset.Robj", v=T)  # glm.fits

# summarize glm fits
fits.sum <- lapply(glm.fits, function(x) return(data.frame(int = x$int,
                                                           pseudo = x$pseudo,
                                                           pval = x$pval)))
jnames <- names(fits.sum)
fits.sum <- bind_rows(fits.sum)
fits.sum$region <- jnames
fits.sum <- fits.sum %>%
  arrange(pseudo, pval)


load("~/Dropbox/scCHiC_figs/FIG4_BM/H3K27me3.datadir_mc_f.Rda", v=T)  # use metacell cluster IDs to color cells in LDA 

# filter cells and peaks 
count.mat <- count.dat$counts
rownames(out.lda@gamma) <- out.lda@documents
cells.keep <- out.lda@documents
peaks.keep <- out.lda@terms
cells.keep.i <- which(colnames(count.mat) %in% cells.keep)
peaks.keep.i <- which(rownames(count.mat) %in% peaks.keep)
count.mat <- count.mat[peaks.keep.i, cells.keep.i]


# Use same coloring as MetaCell output so we can directly compare
Q<-object@mc_fp
mc_index<-object@mc
# mc_colors <- sapply(object@mc, function(x) ifelse(x %in% c(5, 7), "#A9A9A9", "#000000"))
mc_colors<-object@colors
colhash <- hash(seq(length(mc_colors)), mc_colors)
cellhash <- hash(names(mc_index), mc_index)


# summarize glm fits
fits.sum <- lapply(glm.fits, function(x) return(data.frame(int = x$int,
                                                           pseudo = x$pseudo,
                                                           pval = x$pval)))
jnames <- names(fits.sum)
fits.sum <- bind_rows(fits.sum)
fits.sum$region <- jnames
fits.sum <- fits.sum %>%
  arrange(pval)


```

## Comparing MetaCell directly with LDA analysis 

- Metacell shows interesting clusters across cells in H3K27me3

- The cell clusters look like they transition from one stage to another stage (cluster 5 and 7 are the top right, they have high signal in Hox genes)

- If we run LDA on this same count matrix, can we see this transition? Can we use this trajectory to model the counts as a function along this trajectory? This will allow us to find statistically significant windows?

- Cluster 7 shows 

# Results

## PCA on LDA output (gamma matrix) shows some trajectory (PC1, PC2)

```{r lda-compare}


# rownames(out.lda@gamma) <- out.lda@documents
# count.mat <- count.dat$counts
# cells.keep <- out.lda@documents
# peaks.keep <- out.lda@terms
# cells.keep.i <- which(colnames(count.mat) %in% cells.keep)
# peaks.keep.i <- which(rownames(count.mat) %in% peaks.keep)
# count.mat <- count.mat[peaks.keep.i, cells.keep.i]
counts.total <- Matrix::colSums(count.mat)

out.lda.pca <- prcomp(t(out.lda@gamma), center = TRUE, scale. = FALSE)




dat.pca <- as.data.frame(out.lda.pca$rotation) %>%
  mutate(counts.total = counts.total)  # if you want to see if there's bias with library size


rownames(dat.pca) <- colnames(count.mat)

dat.pca$cluster <- sapply(rownames(dat.pca), function(x) cellhash[[x]])
dat.pca$is.5.7 <- sapply(dat.pca$cluster, function(x) ifelse(x %in% c(5, 7), TRUE, FALSE))
dat.pca$is.7 <- sapply(dat.pca$cluster, function(x) ifelse(x %in% c(7), TRUE, FALSE))
dat.pca$is.5 <- sapply(dat.pca$cluster, function(x) ifelse(x %in% c(5), TRUE, FALSE))

mc_colors_vec <- sapply(dat.pca$cluster, function(x) mc_colors[x])
dat.pca$mc_colors <- mc_colors_vec

# ggplot(dat.pca, aes(x = PC1, y = PC2, col = as.character(cluster))) + geom_point(alpha = 0.9)  +
#   theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   ggtitle(paste0("Color by metacell IDs"))

ggplot(dat.pca, aes(x = PC1, y = PC2, col = mc_colors)) + geom_point(alpha = 0.9)  +
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_color_identity() + 
  ggtitle(paste0("Color by metacell IDs"))




```

## PC3 separates between two clusters that were similar in along PC1 and PC2

```{r pc2_pc3}

ggplot(dat.pca, aes(x = PC2, y = PC3, col = mc_colors)) + geom_point(alpha = 0.9)  +
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_color_identity() + 
  ggtitle(paste0("Color by metacell IDs"))

```

## Nonlinear dimensionality reduction captures both a trajectory and separate clusters

```{r umap}

custom.settings <- umap.defaults
custom.settings$n_neighbors <- 20
custom.settings$metric <- "pearson"
custom.settings$n_components <- 2
dat.umap <- umap(out.lda@gamma, config = custom.settings)
rownames(dat.umap$layout) <- rownames(out.lda@gamma)
clstvec <- sapply(rownames(out.lda@gamma), function(x) cellhash[[x]])
is57 <- sapply(clstvec, function(x) ifelse(x %in% c(5, 7), "#A9A9A9", "#000000"))
is7 <- sapply(clstvec, function(x) ifelse(x %in% c(7), "#A9A9A9", "#000000"))
is5 <- sapply(clstvec, function(x) ifelse(x %in% c(5), "#A9A9A9", "#000000"))
plot(dat.umap$layout[, 1], dat.umap$layout[, 2], pch = 20, col = mc_colors_vec)

```

## Analyze PCA1-2 trajectory using principal curve

```{r plot_principal_curve}
proj <- principal_curve(as.matrix(dat.pca[, c("PC1", "PC2")]))

dat.pca.proj <- as.data.frame(proj$s)
dat.pca.proj$lambda <- proj$lambda

ggplot(dat.pca, aes(x = PC1, y = PC2, col = mc_colors)) + geom_point(alpha = 0.9)  +
  theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_color_identity() +
  ggtitle(paste0("Lambda is distance along trajectory. Colors are MetaCell ID")) +
  geom_line(data = dat.pca.proj, inherit.aes = FALSE, aes(x = PC1, y = PC2, alpha = lambda))

```


## Model counts along trajectory using Poisson regression (GLM)


Details: the log counts is a linear function of pseudotime, which is offset by the total counts in each cell (normalize for cells with high counts). Here I show regions with high counts at small pseudotime (decreasing along trajectory).

```{r glm_output}
print(head(fits.sum %>% filter(pseudo < 0)))
```

## Poisson regression model predicts signal at Hox genes decreases along trajectory

It's a naive fitting with lots of details ignored, such as overdispersion, zero-inflation, shrinkage of estimates.

```{r glm_show}

jpeak <- "chr2:74660000-74760000"
fits.again <- FitGlmRow(count.mat[jpeak, ], jpseudo, jsize, returnobj = TRUE)

pred.x <- runif(100, min(jpseudo), max(jpseudo))
jsizenew <- 10000  # just use one cell size for the prediction trend. When fitting, each cell has an offset that accounts for total counts. The trend you see is therefore a cell with 10000 counts. 
newdat <- data.frame(pseudo = pred.x, size = jsizenew)
pred.y <- exp(predict(fits.again, newdata = newdat))

jdat.real <- data.frame(x = jpseudo, y = count.mat[jpeak, ], type = "real")
jdat.pred <- data.frame(x = pred.x, y = pred.y, type = "pred")

ggplot() + geom_point(data = jdat.real, aes(x = x, y = y), alpha=0.15) + 
  geom_line(data = jdat.pred, aes(x = x, y = y)) + theme_bw() + ggtitle(jpeak) + 
  xlab("Length along trajectory") + ylab("Counts")

```


# Conclusions

## Conclusions

- Analysis of count matrix using LDA recapitulates what we see with MetaCell

- Naive modeling of count data along trajectory identifies regions with Hox genes decreasing along trajectory

- At 100kb bins, probably MetaCell is doing better than LDA for analyzing relationships between cells. LDA may be more advantageous as you start to get very sparse matrices. 

## Going further

- Explore the transcriptome and histone marks data so we can see how we should link these datasets together. 

- Model counts along trajectory using a proper noise model. Either a negative binomial for large bins or beta-binomial if we do very small bins.

